Estimation of Spatiotemporal Integro-Difference Equations for Dynamical Neural Field Models

### 2009-11-14 ltx/BrainIDE.tex
added a TODO environment. Use like \todo{Something we should do!}. Have also changed the formatting to draft onecolumn mode so I can sit with it and a coffee and cover it in ink. Which is what I am going to do right now. Saturdays are great!
Fixed typos in equations (3), (5) and (13)
trying to remove files
Attempted to make the model description clearer by adding an alternate derivation

\documentclass[journal]{IEEEtran}

\usepackage{graphicx}
\usepackage{color}                    % For creating coloured text and background
\usepackage{amsmath,amssymb,amsfonts} % Typical maths resource packages
% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

@@

\section{Neural Field Model}
In this section we describe a variant of the  Wilson and Cowan neural field model~\cite{Wilson1973} that will be used to create our connectivity estimator. This model has shown to be descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations and epileptic behavior. It is also capable of generating complex patterns of activity such as Turing patterns, spirals and traveling oscillations. The general form of the model is described here. The post-synaptic potentials generated at a neuronal population $s$ by action potentials arriving from all other connected populations at position $r$ can be described by
\begin{equation}\label{Eq1}
v\left( {s,t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)J\left( {r,t'} \right)dt'}.
\end{equation}
The post-synaptic response kernel $h(t)$ is described by
\begin{equation}\label{Eq2}
h(t) = \left\{ {\begin{array}{*{20}{c}}
   {\exp ( - {t \mathord{\left/
 {\vphantom {t \tau }} \right.
 \kern-\nulldelimiterspace} \tau })} & {t \ge 0}  \\
   0 & {t < 0}  \\
\end{array}} \right.,
\end{equation}
where $\tau$ is the synaptic time constant and typically has values from 10-20~ms (\textcolor[rgb]{1.00,0.00,0.00}{refs}). The synaptic response kernel can be more elaborate, using a rise and fall time constant, however this simple form serves our purposes. Equation~\ref{Eq1} relates average firing rate of a presynaptic neural population to the mean membrane potential of the post-synaptic neurons. The mean firing rate, $J$, is convolved with the response kernel to give membrane potential since the firing rate is assumed to follow a (homogeneous?) Poisson process. Nonlocal interactions between cortical populations are described by	
\begin{equation}\label{Eq3}
J\left( {s,t} \right) = \int_\Omega  {W\left( {s,r} \right)P\left( {v\left( {r,\bar t} \right)} \right)dr}  + E\left( {s,t} \right)
\end{equation}
where $\bar t = t - \frac{{\left| {s - r} \right|}}{d}$, $P$ is a mean pulse rate of population at $y$, $W$ is the connectivity kernel, $E$ is an external input, $\Omega$ is the spatial domain and $d$ is the propagation velocity from $r$ to $s$.

The average firing rate of the presynaptic neurons rate to the pre-synaptic membrane potential by the sigmoid function
\begin{equation}\label{Eq4}
P\left( {v\left( {r,\bar t} \right)} \right) = {R_0}{\left( {1 + \exp \left( {\beta \left( {{v_0}\left( r \right) - v\left( {r,\bar t} \right)} \right)} \right)} \right)^{ - 1}}.
\end{equation}
Even though $v_0$ is a constant, we show express it as a field allowing for inhomogeneities and for clarity in constructing the state-space representation in the next section. The parameters in equation~\ref{Eq4} are... $v_0$ is typically 6~mV, $R_0$ is 5~$s^{-1}$ and $\beta$ is 0.56~mV$^{-1}$ (from Wendling and David papers). By combining equations~\ref{Eq1} and \ref{Eq3} we get the spatiotemporal model
\begin{equation}\label{Eq5}
v\left( {s,t} \right) = \int\limits_{ - \infty }^t {\int\limits_\Omega  {h\left( {t - t'} \right)W\left( {s,r} \right)P\left( {v\left( {r,\bar t} \right)} \right)dy} dt'}  + E\left( {s,t} \right).
\end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{Eq6}
Lh\left( t \right) = \delta \left( t \right),
\end{equation}
where $L$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. This provides the most general form
\begin{equation}\label{Eq7}
L\left( {\frac{\partial }{{dt}}} \right)v\left( {s,t} \right) = \int_\Omega  {W\left( {s,r} \right)P\left( {v\left( {r,\bar t} \right)} \right)dr}  + E\left( {s,t} \right),
\end{equation}
where $L$ is a polynomial of order $n$ with constant coefficients. This provides an $n^{th}$ order model. Most studies considering this model use first or second order time derivatives. We shall use the first order model giving
\begin{equation}\label{Eq8}
\frac{{dv\left( {s,t} \right)}}{{dt}} + \frac{{v\left( {s,t} \right)}}{\tau } = \int_\Omega  {W\left( {s,r} \right)P\left( {v\left( {r,\bar t} \right)} \right)dr}  + E\left( {s,t} \right).
\end{equation}
The neural field model can be represented as a discrete-time stochastic IDE by
\begin{equation}\label{Eq9}
{v_{t + 1}}\left( s \right) = \int_\Omega  {W\left( {s,r} \right)P\left( {{v_{\bar t}}\left( r \right)} \right)dr}  + {\varepsilon _t}\left( s \right) - \frac{1}{\tau }{v_t}\left( s \right),
\end{equation}
where $\epsilon(t)$ is the discrete time stochastic input. The IDE representation of equation~\ref{Eq9} is discrete in time and continuous in space. The neural field model is popular due to being parsimonious yet having a strong link with the underlying physiology.

@@

  + \frac{{{R_0}}}{2}\beta \int_\Omega  {W\left( {\left| {s - r} \right|} \right){v_t}\left( r \right)dr}  \\

>>>>

\documentclass[onecolumn,draftcls]{IEEEtran}

\usepackage{graphicx}
\usepackage{color}                    % For creating coloured text and background
\usepackage{amsmath,amssymb,amsfonts} % Typical maths resource packages
% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}
% TODO environment
\newcommand{\todo}[1]{\textsf{\emph{\textbf{\textcolor{blue}{#1}}}}}

@@

\section{Neural Field Model}
In this section we describe a variant of the  Wilson and Cowan neural field model~\cite{Wilson1973} that will be used to create our connectivity estimator. This model has shown to be descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations and epileptic behavior. It is also capable of generating complex patterns of activity such as Turing patterns, spirals and traveling oscillations. The general form of the model is described here. The post-synaptic potentials generated at a neuronal population $s$ by action potentials arriving from all other connected populations at position $r$ can be described by
\begin{equation}\label{Eq1}
v\left( {s,t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)J\left( {s,t'} \right)dt'}.
\end{equation}
The post-synaptic response kernel $h(t)$ is described by
\begin{equation}\label{Eq2}
h(t) = \left\{ {\begin{array}{*{20}{c}}
   {\exp ( - {t \mathord{\left/
 {\vphantom {t \tau }} \right.
 \kern-\nulldelimiterspace} \tau })} & {t \ge 0}  \\
   0 & {t < 0}  \\
\end{array}} \right.,
\end{equation}
where $\tau$ is the synaptic time constant and typically has values from 10-20~ms (\textcolor[rgb]{1.00,0.00,0.00}{refs}). The synaptic response kernel can be more elaborate, using a rise and fall time constant, however this simple form serves our purposes. Equation~\ref{Eq1} relates average firing rate of a presynaptic neural population to the mean membrane potential of the post-synaptic neurons. The mean firing rate, $J$, is convolved with the response kernel to give membrane potential since the firing rate is assumed to follow a (homogeneous?) Poisson process. Nonlocal interactions between cortical populations are described by	
\begin{equation}\label{Eq3}
J\left( {s,t} \right) = \int_\Omega  {W\left( {s,r} \right)P\left( {v\left( {r,\bar t} \right)} \right)dr}  + E\left( {s,t} \right)
\end{equation}
where $\bar t = t - \frac{{\left| {s - r} \right|}}{d}$, $P$ is a mean pulse rate of population at $y$, $W$ is the connectivity kernel, $E$ is an external input, $\Omega$ is the spatial domain and $d$ is the propagation velocity from $r$ to $s$.

The average firing rate of the presynaptic neurons rate to the pre-synaptic membrane potential by the sigmoid function
\begin{equation}\label{Eq4}
P\left( {v\left( {r,\bar t} \right)} \right) = {R_0}{\left( {1 + \exp \left( {\beta \left( {{v_0}\left( r \right) - v\left( {r,\bar t} \right)} \right)} \right)} \right)^{ - 1}}.
\end{equation}
Even though $v_0$ is a constant, we show express it as a field allowing for inhomogeneities and for clarity in constructing the state-space representation in the next section. The parameters in equation~\ref{Eq4} are... $v_0$ is typically 6~mV, $R_0$ is 5~$s^{-1}$ and $\beta$ is 0.56~mV$^{-1}$ (from Wendling and David papers). By combining equations~\ref{Eq1} and \ref{Eq3} we get the spatiotemporal model
\begin{equation}\label{Eq5}
v\left(s,t\right) = \int\limits_{-\infty}^t\int\limits_\Omega \left[ h\left(t - t'\right)W\left(s,r\right)P\left( v\left( r,\bar t \right)\right)+E\left(s,t'\right)\right] dr dt'
\end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{Eq6}
Lh\left( t \right) = \delta \left( t \right),
\end{equation}
where $L$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. This provides the most general form
\begin{equation}\label{Eq7}
L\left( {\frac{\partial }{{dt}}} \right)v\left( {s,t} \right) = \int_\Omega  {W\left( {s,r} \right)P\left( {v\left( {r,\bar t} \right)} \right)dr}  + E\left( {s,t} \right),
\end{equation}
where $L$ is a polynomial of order $n$ with constant coefficients. This provides an $n^{th}$ order model. Most studies considering this model use first or second order time derivatives. We shall use the first order model giving
\begin{equation}\label{Eq8}
\frac{{dv\left( {s,t} \right)}}{{dt}} + \frac{{v\left( {s,t} \right)}}{\tau } = \int_\Omega  {W\left( {s,r} \right)P\left( {v\left( {r,\bar t} \right)} \right)dr}  + E\left( {s,t} \right).
\end{equation}
The neural field model can be represented as a discrete-time stochastic IDE by
\begin{equation}\label{Eq9}
{v_{t + 1}}\left( s \right) = \int_\Omega  {W\left( {s,r} \right)P\left( {{v_{\bar t}}\left( r \right)} \right)dr}  + {\varepsilon _t}\left( s \right) - \frac{1}{\tau }{v_t}\left( s \right),
\end{equation}
where $\epsilon(t)$ is the discrete time stochastic input. The IDE representation of equation~\ref{Eq9} is discrete in time and continuous in space. The neural field model is popular due to being parsimonious yet having a strong link with the underlying physiology.

\subsection{Alternate model description}

\begin{figure*}
\centering
\includegraphics[scale=0.4]{BasicStructureOfModel.eps}
\caption[NeuralMassModel]{This figure describes the basic interactions of neural masses that constitute the neural field model.}
\label{NeuralMassModel}
\end{figure*}

An alternative description of the model is as follows. The model can be thought of as a continuum of neural masses. To illustrate this idea we shall first describe the interaction of discrete neural masses. Assuming post-synaptic potentials sum linearly, the membrane potential at the soma of neuron $i$ is described as
\begin{equation}\label{Eq1.1}
{v_i}\left( t \right) = \sum\nolimits_{j,k} {{w_{ij}}{h_{j}}\left( t \right)} H\left( {t - {t_k}} \right),
\end{equation}
where $w_{ij}$ is the synaptic weight from neuron $j$ to $i$, $h(t)$ is the synaptic response kernel, $H$ is the Heaviside step function and action potential arrive at neuron I at times $t_k$. Alternatively the membrane potential can described by
\begin{equation}\label{1.2}
{v_i}\left( t \right) = \sum\nolimits_j {\int_{ - \infty }^t {{w_{ij}}{h_j}\left( {t - t'} \right){r_j}\left( t \right)} } dt',
\end{equation}
where $r_j(t)$ is the firing rates of the presynaptic neurons. To convert the equation system into and integral differential system we shall take derivative of~\ref{Eq1.1} yielding
\begin{eqnarray}
% \nonumber to remove numbering (before each equation)
  \frac{{d{v_i}\left( t \right)}}{{dt}} &=& \sum\nolimits_{j,k} {\left( {{w_{ij}}{h_j}\left( t \right)\frac{d}{{dt}}H\left( {t - {t_k}} \right) + {w_{ij}}H\left( {t - {t_k}} \right)\frac{d}{{dt}}{h_j}\left( t \right)} \right)} \nonumber \\
   &=& \sum\nolimits_{j,k} {{w_{ij}}{h_j}\left( t \right)\delta \left( {t - {t_k}} \right)}  - \sum\nolimits_{j,k} {{w_{ij}}H\left( {t - {t_k}} \right)\frac{1}{{{\tau _j}}}{h_j}\left( t \right)}   \nonumber \\
   &=& \sum\nolimits_{j,k} {{w_{ij}}{h_j}\left( t \right)\delta \left( {t - {t_k}} \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\frac{1}{{{\tau _j}}}} \nonumber  \\
   &=& \sum\nolimits_j {{w_{ij}}{r_j}\left( t \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\frac{1}{{{\tau _j}}}} \nonumber \\
   &=& \sum\nolimits_j {{w_{ij}}S\left( {{v_j}\left( t \right)} \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\frac{1}{{{\tau _j}}}}. 
\end{eqnarray}
This form can easily be extended to represent a continuous field.

@@

  + \frac{{{R_0}}}{4}\beta \int_\Omega  {W\left( {\left| {s - r} \right|} \right){v_t}\left( r \right)dr}  \\
